---
title: "msqrob2PTM: msqrob2 Mass Spectrometry Data Analysis Workflow for differential PTM analysis - DIA-NN case study"
author:
    - name: Lieven Clement, Nina Demeulemeester, Luca De Corso and Lucas Beerland
output: 
    BiocStyle::html_document:
      code_download: true
      toc: true
      toc_float: true
      number_sections: true

date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('msqrob2')`"
vignette: >
    %\VignetteIndexEntry{msqrob2PTM use case}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---


# Background 

Mutations in tRNAs can lead to mis-incorporation of an amino acid into a growing 
polypeptide chain that differs from what is specified by the mRNA in a process 
known as mistranslation. 
To better understand the impact of mistranslating tRNA variants, Berg et al. 
profiled the proteome and phosphoproteome of yeast expressing three different 
mistranslating tRNAs. 
The first is a proline tRNA with a G3:U70 base pair in its acceptor stem that 
mis-incorporates alanine at proline codons (Pro -> Ala; Hoffman et al. 2017). 
The other two are serine tRNAs with either a UGG proline or UCU arginine 
anticodon which mis-incorporate serine at proline (Pro -> Ser) or serine at 
arginine codons (Arg -> Ser), respectively (Berg et al. 2017, 2019b). 

Six replicates of each strain were grown. From each replicate an non-enriched (whole proteome)
and a phospopeptide - enriched sample was prepared and quantified using a data
independent acquisition workflow.

The data for the non-enriched and phospho enriched samples are deposited to 
PRIDE, dataset identifiers PXD068388 and PXD068392). 

# Load packages

We load the `msqrob2` package, along with additional packages for
data manipulation and visualisation.

```{r load_libraries}
library("QFeatures")  
library("dplyr") 
library("tidyr")
library("ggplot2")
library("msqrob2")    
library("stringr")
library("ExploreModelMatrix")
library("MsCoreUtils") 
library("matrixStats")
library("patchwork")
library("kableExtra")
library("ComplexHeatmap")
library("purrr")
library("tibble")
library("scater")
library("BiocFileCache")
```

# Data

## Precursor table

We load the output from Spectronaut parquet file. Can be file path to local file or url to file that lives on the web.

```{r data_path}
bfc <- BiocFileCache("/opt/fileCache")
precursorFile <- bfcrpath(bfc,"https://zenodo.org/records/20414816/files/WholeProteome_DIANNreport.parquet?download=1")
precursorFilePTM <- bfcrpath(bfc,"https://zenodo.org/records/20414816/files/Phosphoproteome_DIANNreport.parquet?download=1")
```


We can import the report.parquet file using the `read_parquet` function from the `arrow` package. 

```{r data_import}
precursors <- arrow::read_parquet(precursorFile) # function from the arrow package
precursorsPTM <- arrow::read_parquet(precursorFilePTM) # function from the arrow package
#precursors <- data.table::fread(precursorFile) # For older versions the results are stored as tsv files. 
```

We subset the precursor files using only the variables that we need. Note that 
`Protein.Sites` refers to the exact modified residue position in the full 
protein sequence, which will be useful when assessing PTM-level quantification. 
Note, that we filter out the strain with the ProSer variant here. 

```{r data_subset}
precursors <- precursors |> 
  filter(!grepl("ProSer",Run)) |>
  select(
         Run, 
         Precursor.Id, 
         Modified.Sequence, 
         Stripped.Sequence, 
         Precursor.Charge, 
         Protein.Group, 
         Protein.Names,
         Protein.Ids,
         Protein.Sites, #exact modified residue position in the full protein sequence
         Genes,
         Precursor.Quantity,
         Precursor.Normalised,
         Normalisation.Factor,
         Ms1.Area,
         Ms1.Normalised,
         PG.MaxLFQ,
         Q.Value, 
         Lib.Q.Value,
         PG.Q.Value,
         Lib.PG.Q.Value,
         Proteotypic,
         Decoy, # Not available in older versions of DIA-NN
         RT)
precursorsPTM <- precursorsPTM |> 
  filter(!grepl("ProSer",Run)) |>
  select(
         Run, 
         Precursor.Id, 
         Modified.Sequence, 
         Stripped.Sequence, 
         Precursor.Charge, 
         Protein.Group, 
         Protein.Names,
         Protein.Ids,
         Protein.Sites, #exact modified residue position in the full protein sequence
         Genes,
         Precursor.Quantity,
         Precursor.Normalised,
         Normalisation.Factor,
         Ms1.Area,
         Ms1.Normalised,
         PG.MaxLFQ,
         Q.Value, 
         Lib.Q.Value,
         PG.Q.Value,
         Lib.PG.Q.Value,
         Proteotypic,
         Decoy, # Not available in older versions of DIA-NN
         RT)
```


## Subset phospho

In DIA-NN phosphorylation is indicated with (UniMod:21) in variable 
`Modified.Sequence`, which contains the precursor sequence along with its 
modifications. 

Here, we only keep precursors that are phosphorylated. 

```{r}
precursorsPTM <- precursorsPTM |>
  filter(grepl(pattern = "\\(UniMod:21\\)", Modified.Sequence))
```

## Quick check for imputation

Quick check on distribution of precursors MS2 intensities.

```{r quick check}
precursors |> 
  filter(is.finite(Precursor.Quantity) & Precursor.Quantity > 0) |>
  ggplot(aes(x = log2(Precursor.Quantity))) +
  geom_density() +
  theme_minimal() +
  labs(subtitle = "Abuandances (non-enriched)") + precursorsPTM |> 
  filter(is.finite(Precursor.Quantity) & Precursor.Quantity > 0) |>
  ggplot(aes(x = log2(Precursor.Quantity))) +
  geom_density() +
  theme_minimal() +
  labs(subtitle = "Abundances (phospho enriched)")
```


Seems no imputation has been done. 

## Sample annotation table

The [sample annotation table](#sec-annotation_table)) is not available
and can be generated from the run labels, as the researchers included information on the design in the filenames.

1. Extract unique Run labels
2. Split Run label into variables using the "_" separator
3. Rename the variable Run to runCol (needed for `readQFeatures` function)
4. Add sampleId
5. Sort annotation file according to sampleId

```{r metadata_non_enriched}
annot <- precursors |> 
  dplyr::distinct(Run) |>  ## 1.
  separate(Run,
           into=c("run","label2","strain","rep","type","acquisition"),
           sep="_",
           remove=FALSE) |>   ## 2.
  rename(runCol = Run) |> ## 3.
  mutate(sampleId = paste0(strain,rep)) |> #4.
  arrange(sampleId) #5.
annot
```

We do the same for the phospho data.

```{r meta_data_phospho}
annotPTM <- precursorsPTM |> 
   dplyr::distinct(Run) |>  ## 1.
  separate(Run,
           into=c("label1","label2","strain","rep","type","acquisition"),
           sep="_",
           remove=FALSE) |>   ## 2.
  rename(runCol = Run) |> ## 3.
  mutate(sampleId = paste0(strain,rep)) |> #4.
  arrange(sampleId) #5.
annotPTM
```

## Convert to QFeatures

First, recall that the precursor table is file in long format. 
Every quantitative column in the precursor table contains
information for multiple runs. Therefore, the function split the table
based on the run identifier, given by the `runCol` argument (for
Spectronaut, that identifier is contained in `run`). 
So, the `QFeatures` object after import will contain as many sets as there are
runs. 
Next, the function links the annotation table with the PSM data.
To achieve this, the annotation table must contain a `runCol` column
that provides the run identifier in which each sample has been
acquired, and this information will be used to match the identifiers
in the `Run` column of the precursor table. 

Here, we will use the `Precursor.Quantity` column as quantification input. 

1. We first combine the precursor and precursorPTM table in a list of 
QFeatures objects, 
2. which we subsequently reduce in one QFeatures object. 

```{r import_qfeatures}
(qf <- list(
  readQFeatures(assayData = precursors,  
                              colData = annot,
                              quantCols = "Precursor.Quantity",
                              runCol = "Run",
                              fnames = "Precursor.Id"),
  readQFeatures(assayData = precursorsPTM,  
                              colData = annotPTM,
                              quantCols = "Precursor.Quantity",
                              runCol = "Run",
                              fnames = "Precursor.Id") ##1.
) |> purrr::reduce(c) ##2.
)
```


# Data preprocessing{#sec-basic_preprocess}

Here we conduct the following processing steps

1. Replace zero's by NA.
2. Precursor filtering and assay joining
3. Log-transformation
4. Normalisation
4. Summarisation for non-enriched runs 

## Encoding missing values

We first replace any zero in the quantitative data
with an `NA`. 

```{r zero_to_na}
qf <- zeroIsNA(qf, names(qf))
```

Note that `msqrob2` can handle missing data without having to rely on
hard-to-verify imputation assumptions, which is our general recommendation. However, `msqrob2` does not
prevent users from using imputation, which can be performed with
`impute()` from the `QFeatures` package. 

## Precursor Filtering 

Filtering removes low-quality and unreliable precursors that would otherwise 
introduce noise and artefacts in the data. 

### Remove questionable identifications

We apply standard filtering: 


1. q-value threshold of 0.01 for the identification of precursors (`Q.Value`) and protein groups (`PG.Q.Value`). If the file is processed via matching between runs, it is also useful to filter on Lib.Q.Value and Lib.PG.Q.Value.

2. Remove precursors that could not be mapped, i.e. when `Precursor.Id` column is an empty string.  

3. Filter decoys, i.e. only keep precursors for which the `Decoy` column equals 0. (Note, that the `Decoy` column is not present in the output of older versions of DIA-NN)

4. Keeping only proteotypic peptides, which map uniquely to a specific protein.



```{r filtering}
qf <-  qf |>
  filterFeatures(~ Q.Value <= 0.01 & #1.
                     PG.Q.Value <= 0.01 &  #1.
                     Lib.Q.Value <= 0.01 & #1.
                     Lib.PG.Q.Value <= 0.01 & #1.
                     Precursor.Id != "" & #2.
                     Decoy == 0 & #3. (Note, that this filter is not available with previous versions of DIA-NN, as the report.tsv file did not include a Decoy column. So other strategies are needed if Decoys are in the output file.)
                     Proteotypic == 1) #4.
```


Note, that it is important that the filtering criteria are not distorting the distribution of the test statistics in the downstream analysis for features that are non-DA. 
It can be shown that filtering will not induce bias results when the filtering criterion is independent of test statistic. The criteria that we proposed above are all based on the results of the identification step, hence, they are independent of the downstream test statistics that will be used to prioritize DA proteins. 

### Assay joining

Up to now, the data from different runs were kept in separate assays. We can 
now join the normalised sets into an precursor set using joinAssays(). Sets are 
joined by stacking the columns (samples) in a matrix and rows (features) are 
matched according to a row identifier, here the `Precursor.Id`.
We do this for the non-enriched and phospho samples, which we store in assays 
with names: `precursors` and `precursorsPTM`, respectively.

```{r assay_joining}
qf <- joinAssays(
  x = qf,
  i = grep("Proteome",names(qf), value = TRUE),
  fcol = "Precursor.Id",
  name = "precursors"
  )
(qf <- joinAssays(
  x = qf,
  i = grep("Phospho",names(qf), value = TRUE),
  fcol = "Precursor.Id",
  name = "precursorsPTM"
  )
)
```

### Filtering: Remove highly missing precursors

We keep peptides that were observed at last 6 times out of the $n
= 24$ samples, so that we can estimate the peptide characteristics. 

```{r remove_precursors_with_missingness}
nObs <- 6
n <- ncol(qf[["precursors"]])

(qf <- filterNA(qf, i = c("precursors", "precursorsPTM"), pNA = (n - nObs) / n))
```


### Filter one-hit wonders

Here, we remove proteins that can only be found by one peptide, as such proteins 
may not be trustworthy. We only do that for the non-enriched samples as we will
use this assay to estimate protein abundances.

1. We first calculate how many distinct peptides map to each protein group (`PG_ProteinGroups`). We use the stripped precursor sequence, i.e. sequence of the base peptide for this purpose. 

2. We store this information in the row data of the precursors assay

3. We filter precursors of one-hit wonder proteins.  


```{r filter_one_hit_wonders}
# Filter for peptides per protein
pepsPerProtDf <- qf[["precursors"]] |> 
    rowData() |> 
    data.frame() |>
    dplyr::select("Stripped.Sequence", "Protein.Group") |>
    group_by(Protein.Group) |>
    mutate(pepsPerProt = Stripped.Sequence |>
               unique() |> length()
           ) #1.

rowData(qf[["precursors"]])$pepsPerProt <- pepsPerProtDf$pepsPerProt #2. 

qf <- filterFeatures(qf, 
                              ~ pepsPerProt > 1, 
                              keep = TRUE) #3.
```

## Log-transformation 


We perform log2-transformation with `logTransform()` from the
`QFeatures` package. We use `base = 2` and store the result in a new 
summarized experiment named `precursors_log`.

```{r log_transformation}
qf <- logTransform(qf, 
                            base = 2, 
                            i = "precursors", 
                            name = "precursors_log")
qf <- logTransform(qf, 
                            base = 2, 
                            i = "precursorsPTM", 
                            name = "precursorsPTM_log")
```

## Normalisation {#sec-norm}

We first evaluate the distributions of the non-normalised precursors per sample.

```{r}
qf[, , "precursors_log"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "precursors") +
  theme_minimal()
  
qf[, , "precursorsPTM_log"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "PTM - precursors") +
  theme_minimal()
```

The data seems very clean. 

However, normalisation is still required. 
For this dataset we use median of ratio normalisation, which addresses both 
differences in loading as well as difference in compositionality.
We first calculate a reference sample by using the log2 geometric mean, i.e. 
the arithmetic mean on a log scale, for every feature.
Subsequently log2 fold changes are calculated between each feature of a sample 
and the reference sample.
Finally, the median of these log2 fold changes are used as the log2-tranformed 
normalisation factor. 
The method is the default method for normalisation in bulk RNA-seq data analysis 
with DESeq2.
In our hands it performs very well in a proteomics context.

```{r normalisation}
qf <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf,
  MARGIN = 2,
  STATS = nfLogMedianOfRatios(qf,"precursors_log"),
  i = "precursors_log",
  name = "precursors_norm"
)
qf <- sweep( #4. Subtract log2 norm factor column-by-column (MARGIN = 2)
  qf,
  MARGIN = 2,
  STATS = nfLogMedianOfRatios(qf,"precursorsPTM_log"),
  i = "precursorsPTM_log",
  name = "precursorsPTM_norm"
)
```


We explore the effect of the global normalisation in the subsequent plot. 

Formally, the function applies the following operation on each sample
$i$ across all precursors $p$:

$$
y_{ip}^{\text{norm}} = y_{ip} - \log_2(nf_i)
$$

with $y_ip$ the log2-transformed intensities and $nf_i$ the log2-transformed norm factor. 
Upon normalisation, we can see that the distribution of the 
$y_{ip}^{\text{norm}}$ nicely overlap (using the same code as above)

```{r assess_normalisation_all}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "Median normalisation precursors") +
  theme_minimal()
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "Median normalisation PTM precursors") +
  theme_minimal()
```

We also explore the normalisation for the common precursors. 


```{r assess_normalisation_common}
precursorsCommon <- assay(qf[, , "precursors_log"]) |> na.exclude() |> rownames()
precursorCommonDens <- qf[precursorsCommon, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Median normalised log2 precursor intensities (common)")
precursorCommonDens
precursorsCommonPTM <- assay(qf[, , "precursorsPTM_log"]) |> na.exclude() |> rownames()
precursorCommonDensPTM <- qf[precursorsCommonPTM, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Median normalised log2 precursor intensities (common)")
precursorCommonDensPTM
```

## Summarisation

Here, we summarise the precursor-level data for the non-enriched runs into 
protein intensities using maxLFQ. 
maxLFQ first calculates all pairwise log ratio's between samples only using 
their shared precursors.
Particularly, it uses the median of the log ratio's between the shared precursors s, i.e. 
$$r_{ij} = median(y_{sj}-y_{si})$$.  
Hence, it first eliminates ion effects. 
It then estimates the summaries by solving 

$$
\sum_i\sum_j(y^{prot}_j - y^{prot}_i - r_ij)^2
$$
It is implemented in the `maxLFQ` function of the `iq` package. 

`aggregateFeatures()` streamlines summarisation. It requires the name
of a `rowData` column to group the precursors into proteins (or
protein groups), here `PG_ProteinGroups`. We provide the summarisation
approach through the `fun` argument. Other summarisation methods
are available from the `MsCoreUtils` package, see `?aggregateFeatures`
for a comprehensive list. The function will return a `QFeatures`
object with a new set that we called `proteins`.

```{r summarisation, warning=FALSE}
(qf <- aggregateFeatures(
  qf, i = "precursors_norm", 
  name = "proteins",
  fcol = "Protein.Group", 
  fun = function(X) iq::maxLFQ(X)$estimate
))
```


# Data exploration and QC 

Data exploration aims to highlight the main sources of variation in
the data prior to data modelling and can pinpoint to outlying or
off-behaving samples. 

## Phospho-enriched samples 

### Marginal distribution at precursor level 


```{r qc_ptm_density}
precursorDensPTM <- qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 PTM precursor intensities")
precursorDensPTM
```

```{r qc_ptm_boxplot}
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 PTM precursor intensities")
```


### Charge state 

```{r qc_ptm_chargestate}
qf[, , "precursorsPTM_norm"] |>
  longForm(colvars = colnames(colData(qf)), rowvars = "Precursor.Charge") |>
  as.data.frame() |>
  filter(!is.na(value)) |>
  filter(Precursor.Charge<=4) |> 
ggplot(aes(x = sampleId)) +
  geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
           colour = "black") +
  labs(title = "Number of phospho-ions per sample",
       x = "Sample",
       fill = "Charge state") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```

## Identifications per sample 

```{r qc_ptm_id}
qf[,,"precursorsPTM_norm"] |> 
  longForm(colvars = colnames(colData(qf)), 
           rowvars= c("Precursor.Id", 
                      "Protein.Group")) |>
  data.frame() |>
  filter(!is.na(value)) |>
  group_by(strain, sampleId) |>
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) |> 
  pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = sampleId, y = IDs, fill = strain)) +
  geom_col() +
  #scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per PTM sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
# Data Modeling (Robust Regression){#sec-modelling}
```

## Dimensionality reduction plot 

A common approach for data exploration is to
perform dimension reduction, such as Multi Dimensional Scaling (MDS).
We will first extract the set to explore along the sample annotations
(used for plot colouring).

```{r qc_ptm_mds}
precursorPTMMDS <- getWithColData(qf, "precursorsPTM_norm") |> 
  as("SingleCellExperiment") |> 
  runMDS(exprs_values = 1) |> 
  plotMDS(colour_by = "strain") 
precursorPTMMDS
```

## Non-enriched samples

### Marginal distribution at precursor and protein level 

```{r qc_ne_density}
precursorDens <- qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 precursor intensities")
precursorDens
```

```{r qc_ne_boxplot}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "Normalised log2 precursor intensities")
```

```{r qc_ne_prot_density}
protDens <- qf[, , "proteins"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  theme_minimal() +
   labs(subtitle = "log2 protein intensities (normalised precursors)")
protDens
```

```{r qc_ne_prot_boxplot}
qf[, , "proteins"] |>
  longForm(colvars = colnames(colData(qf))) |> 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = sampleId,
      y = value,
      colour = strain,
      group = colname) +
  xlab("sample") +
  geom_boxplot() +
  theme_minimal() +
   labs(subtitle = "log2 protein intensities (normalised precursors)")
```


### Charge state 

```{r qc_ne_chargestate}
qf[, , "precursors_norm"] |>
  longForm(colvars = colnames(colData(qf)), rowvars = "Precursor.Charge") |>
  as.data.frame() |>
  filter(!is.na(value)) |>
  filter(Precursor.Charge<=4) |> 
ggplot(aes(x = sampleId)) +
  geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
           colour = "black") +
  labs(title = "Number of precursor ions per sample",
       x = "Sample",
       fill = "Charge state") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

### Identifications per sample 

```{r qc_ne_id}
qf[,,"precursors_norm"] |> 
  longForm(colvars = colnames(colData(qf)), 
           rowvars= c("Precursor.Id", 
                      "Protein.Group")) |>
  data.frame() |>
  filter(!is.na(value)) |>
  group_by(strain, sampleId) |>
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) |> 
  pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = sampleId, y = IDs, fill = strain)) +
  geom_col() +
  #scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))

```

## Dimensionality reduction plot 

A common approach for data exploration is to
perform dimension reduction, such as Multi Dimensional Scaling (MDS).
We will first extract the set to explore along the sample annotations
(used for plot colouring).

```{r qc_ne_prot_mds}
protMDS <- getWithColData(qf, "proteins") |> 
  as("SingleCellExperiment") |> 
  runMDS(exprs_values = 1) |> 
  plotMDS(colour_by = "strain") 
protMDS
```







# Data Modeling at phospho-peptidoform-level 

## Model estimation

1. We first define the model. We only have one source of variability in the 
experiment that we can model, i.e. the effect of the strain. 

2. We fit the model to each phosphorylated precursor in the `precursorsPTM_norm` 
summarised experiment of the QFeatures object `qf` using the msqrob function. 

```{r model, warning=FALSE}
model <- ~ strain
qf <- msqrob(
  qf,
  i = "precursorsPTM_norm",
  formula = model,
  robust = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

## Inference

We can now convert the research question "do the spike-in
conditions affect the protein intensities?" into a statistical
hypothesis. 

In other words, we need to translate this question in a null and alternative hypothesis on a single model parameter or a linear combination of model parameters, which is also referred to with a [contrast](#sec-inference).

To aid defining contrasts, we will visualise the experimental design using the `ExploreModelMatrix` package.

```{r design}

vd <- ExploreModelMatrix::VisualizeDesign(
    sampleData =  colData(qf),
    designFormula = model,
    textSizeFitted = 4
)
vd$plotlist
```
We have 2 interesting research questions related to the diet. 

1. Is there an effect of diet in male mice / prioritise proteins that are DA between low and high diet for male mice
2. Is there an effect of sex in high diet 

Each of these contrasts can be translated in a log2 FC or a difference in log2 FC and can be estimated with one or a linear combination of the model parameters. 

Below we define the contrast and construct the corresponding contrast matrix. 

We can generate the null-hypothesess for all pairwise comparisons manually.
However, here we will use the `createPairwiseContrasts` function in `msqrob2`.

```{r define_H0}
(allHypotheses <- createPairwiseContrasts(
  model, colData(qf), "strain"
  )
)
```

Based on these null hypothesis we now generate the contrast matrix with all
three contrasts using the `makeContrast` function. Note, that experienced users
can also define the contrast matrix, directly. 

```{r make_contrasts}
(L <- makeContrast(
  allHypotheses,
  parameterNames = colnames(vd$designmatrix)
))
```


We assess the contrast for each precursor. 

```{r hypothesis_tests, warning=FALSE}
qf <- hypothesisTest(qf, i = "precursorsPTM_norm", contrast = L, overwrite = TRUE)
```

We extract the results table from the `precursorsPTM_norm` summarised experiment 
in the `qf` object. 

```{r collect_results_PTM}
inferencesPTM <- 
  msqrobCollect(qf[["precursorsPTM_norm"]], L)
```

## Report results

We report the results using a results table, volcano plots and heatmaps. 

### Results table

We use the 5% nominal FDR level to report results 
```{r}
alpha <- 0.05
```

```{r inference_tables_PTM}
#| results: asis
for (j in colnames(L)) {

  inference <- inferencesPTM |> 
    dplyr::filter(adjPval < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant precursors)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

### Volcanoplots 

```{r}
volcanoplotPTM <- plotVolcano(inferencesPTM) + 
  facet_wrap(~contrast) + 
  labs(title = "Precursor-level Phospho DA")
volcanoplotPTM
```



```{r}
inferencesPTM |> 
  filter(adjPval < alpha) |> 
  mutate(DA = sign(logFC) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

### Heatmaps

We will make a heatmap for each contrast. 
Note that we cluster the rows ourselves to avoid issues with missing values. 

```{r precursor_da_heatmaps}
heatmaps <- lapply(colnames(L),
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- se$sampleId #specific to this dataset to get short colnames
                     rowclushlp <- quants
                     rowclushlp[is.na(rowclushlp)] <- min(quants,na.rm=TRUE) - 2
                     rowclus <- hclust(dist(rowclushlp))
                     annotations <- columnAnnotation(
                       group = se$strain
                       ) #3.
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0"),
                       cluster_rows = rowclus
                       )
                     )
                   },
                   se = getWithColData(qf, "precursorsPTM_norm"),
                   alpha = alpha)
heatmaps
```



### Detail plots

We can explore the data for a protein to validate the statistical inference results. For example, let’s explore the normalised precursor and the summarised protein intensities for the protein with the most significant log2 fold change.

```{r most_siginificant_PTM}
(target_feature <- inferencesPTM |> 
   dplyr::slice(which.min(pval)) |> 
   pull(feature)
)
inferencesPTM |> 
  filter(feature == target_feature)
```

To obtain the required data, we perform a little data manipulation pipeline:

We use the QFeatures subsetting functionality to retrieve all data related to 
ion data used for model fitting. We then convert the data with longForm() 
for plotting. Finally, we plot the log2 normalised intensities for each sample 
for the unnormalised and normalised phospho precursors. 

```{r detail_line_plot}
qf[target_feature, , c("precursorsPTM_log","precursorsPTM_norm")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |> 
  ggplot() +
  aes(x = sampleId,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r detail_boxplot}
qf[target_feature, , c("precursorsPTM_log","precursorsPTM_norm")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |>
  filter(!is.na(value)) %>% 
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

We prioritised differentially abundant phospho-precursors. However, this might 
be due to the parent proteins on which the PTMs occur also change in abundance 
regardless of the PTM. 
Any changes in the abundance of a phospho-peptidoform can thus be confounded 
with changes in protein abundance.

We therefore assess differential abundance at the protein level for the 
non-enriched samples.

# Data Modeling at protein-level (non-enriched samples)

## Model estimation

We can use the same model to model the protein-level data of the non-enriched 
runs. Indeed, they stem from the same biological samples. 

```{r model_ne, warning=FALSE}
qf <- msqrob(
  qf,
  i = "proteins",
  formula = model,
  robust = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

## Inference

We assess the contrast for each protein. 

```{r inference_ne, warning=FALSE}
qf <- hypothesisTest(qf, i = "proteins", contrast = L, overwrite = TRUE)
```

We extract the results table from the `proteins` summarised experiment in the `qf` object. 

```{r results_ne} 
inferences <- 
  msqrobCollect(qf[["proteins"]], L)
```

## Report results

We report the results using a results table, volcano plots and heatmaps. 

### Results table


```{r result_tables_ne}
#| results: asis
for (j in colnames(L)) {

  inference <- inferences |> 
    dplyr::filter(adjPval < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant proteins)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

### Volcanoplots 

```{r volcanoplots_ne}
volcanoplot <- plotVolcano(inferences) + 
  facet_wrap(~contrast) + 
  labs(title = "Non-enriched")
volcanoplot
```


```{r sigres_ne}
inferences |> 
  filter(adjPval < alpha) |> 
  mutate(DA = sign(logFC) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

We notice that many proteins are DA at the nominal 5% FDR-level. 

### Heatmaps

```{r precursor_da_heatmaps_ne}
heatmaps <- lapply(colnames(L),
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- se$sampleId #specific to this dataset to get short colnames
                     rowclushlp <- quants
                     rowclushlp[is.na(rowclushlp)] <- min(quants,na.rm=TRUE) - 2
                     rowclus <- hclust(dist(rowclushlp))
                     annotations <- columnAnnotation(
                       group = se$strain
                       ) #3.
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0"),
                       cluster_rows = rowclus
                       )
                     )
                   },
                   se = getWithColData(qf, "proteins"),
                   alpha = alpha)
heatmaps
```

### Detail plots

We can explore the data for a protein to validate the statistical inference results. For example, let’s explore the normalised precursor and the summarised protein intensities for the protein with the most significant log2 fold change.

```{r}
(target_feature <- inferences |> 
   dplyr::slice(which.min(pval)) |> 
   pull(feature)
)
inferences |> 
  filter(feature == target_feature)
```

To obtain the required data, we perform a little data manipulation pipeline:

We use the QFeatures subsetting functionality to retrieve all data related to and focusing on the peptides_log and proteins sets that contains the peptide ion data used for model fitting. We then convert the data with longForm() for plotting. Finally, we plot the log2 normalised intensities for each sample at the protein and at the peptide level. Since multiple peptides are recorded for the protein, we link peptides across samples using a grey line. Samples are colored according to E. coli spike-in condition.

```{r}
qf[target_feature, , c("precursors_log","precursors_norm", "proteins")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |> 
  ggplot() +
  aes(x = colname,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r}
qf[target_feature, , c("precursors_log","precursors_norm", "proteins")] |> #1
  longForm(colvars = colnames(colData(qf))) |> #2
  data.frame() |>
  filter(!is.na(value)) %>% 
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

### Compare phospo-precursor level intensities to protein level intensities
 
We compare the intensities for the top 5 DA phospo-precursors for the first contrast to these of their corresponding protein in the non-enriched assay. 
 
```{r comparison_plots, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesPTM |> 
  filter(contrast == contr) |> 
  arrange(pval) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) 
 print(comparison_plot)
}
```

# Data Modeling of phospho-peptidoform usages

## Background 

- Differential abundant PTMs can be due protein abundance

- Port concept of differential usage towards PTMs: 

$$\log_2 DU = \log_2 FC^\text{PTM} - log_2 FC^\text{protein}$$

Two designs: PTM and protein level quantification be done 

- on same sample: paired design 
- using independent samples


### Paired samples 

1. Perform regular data processing at the protein-level: 
  
    - log2-transformation, 
    - normalisation and 
    - summarisation

2. Calculate relative log2 transformed abundances at peptidoform level 
$$y^\text{pep,rel}_i = y^\text{pep}_i - y^\text{prot}_i$$

3. Perform conventional differential msqrob2 analysis on the relative peptidoform level. $\rightarrow$ Immediately gives us diferential usages!
    
4. Optionally 
    
  a. aggregate relative log2 transformed abundances at peptidoform level at PTM-level
  b. perform a differential msqrob2 analysis at PTM-level


### Unpaired PTM-level and protein samples 

This is the default MSstatsTMT approach, which they also use for paired designs?!

  1. Perform regular msqrob2 differential analysis at the protein-level
  2. Perform regular msqrob2 differential analysis at the peptidoform (PTM-level)
  3. Calculate usages
  
$$ 
\begin{array}{ccc}
\log_2 \widehat{DU} & =& \log_2 \widehat{FC}^\text{PTM} - log_2 \widehat{FC}^\text{protein}\\
SE_\widehat{DU} &=& \sqrt{(SE_\widehat{FC}^{PTM})^2 + (SE_\widehat{FC}^{protein})^2}
\end{array}
$$
### Our study

For this dataset we have phospho-enriched and non-enriched runs on the same 
biological samples. So we have a paired design! 


## Paired samples: calculate usages

We can calculate the usages for every precursor ion by subtracting the 
corresponding log-transformed protein level abundance values from the log2 
transformed and normalized precursor ion intensities. 

We first develop a function that will extract the corresponding protein level 
abundance values. 

It has following inputs: 

- `se`: summarised experiment of the log2-transformed and normalized precursor ion 
- `seNE`: summarised experiment of the log2-transformed, normalized and aggregated 
  protein level abundances. 
- `fcol`: annotation column with protein name for the precursor ions
- `fcolNE`: annotation column with protein name for the protein level abundances
- `matchColsSeToseNE`: a vector that matches the columns of `seNE` to these in 
  `se`. 

The function 

1. initiates the matrix with log2 tranformed correction factors at NA. 
2. Matches the rows of se with these of seNE using the annotation columns
3. Extracts the protein abundance values for the precursors from the matching 
  rows and columns in the protein level assay.
  
```{r}
calculate_usage_nfLog <- function(se, 
                               seNE, 
                               fcol = "Protein.Group", 
                               fcolNE = "Protein.Group", 
                               matchColsSeToseNE = 1:ncol(se)
                               )
{
    nf <- assay(se); nf[,] <- NA #1. 
    matchSeRowsToSeNERows <- match(
      rowData(se)[[fcol]], 
      rowData(seNE)[[fcolNE]]) #2.
    idsMatch <- which(!is.na(matchSeRowsToSeNERows))
    nf[idsMatch, ] <- assay(seNE)[matchSeRowsToSeNERows[idsMatch], matchColsSeToseNE] #3.
    return(nf)
}
```

Here we 

1. match the samples of the phosho enriched samples to these of the 
non-enriched assay. 
2. calculate the correction factors 
3. make a new assay with usages by sweeping out the log2 protein abundances from 
   from the log2 precursor abundances.   
   
```{r}
matchedSamps <- match(
  getWithColData(qf,"precursorsPTM_norm")$sampleId,
  getWithColData(qf,"proteins")$sampleId
  )

nfLog <- calculate_usage_nfLog(qf[["precursorsPTM_norm"]], 
                               qf[["proteins"]], 
                               fcol= "Protein.Group", 
                               fcolNE = "Protein.Group", 
                               matchColsSeToseNE = matchedSamps)

qf <- sweep( #4. Subtract log2 norm factor elementwise (MARGIN = 1:2)
      qf,
      MARGIN = 1:2,
      STATS = nfLog,
      i = "precursorsPTM_norm",
      name = "precursorsPTM_usage"
      )
```

We only keep precursors for which the usage can be calculated in at least 6 samples. 

```{r}
n <- ncol(qf[["precursorsPTM_usage"]])
qf <- filterNA(qf, i = "precursorsPTM_usage", pNA = (n - nObs) / n)
```

We assess the distribution of the usages. 

```{r}
qf[, , "precursorsPTM_usage"] |> #1.
  longForm(colvars = colnames(colData(qf))) |>  #2. 
  data.frame() |> 
  filter(!is.na(value)) |>
  ggplot() + #3.
  aes(x = value,
      colour = strain,
      group = colname) +
  geom_density() +
  labs(subtitle = "Usage") +
  theme_minimal()
```


## Model estimation

Again we can estimate the same model but now for the usages. 

```{r model_usage, warning=FALSE}
qf <- msqrob(
  qf,
  i = "precursorsPTM_usage",
  formula = model,
  robust = TRUE,
  overwrite = TRUE)
```

We enabled M-estimation (`robust = TRUE`) for improved robustness
against outliers. 

## Inference

Contrasts remain the same. 
We assess the contrast for each precursor. 

```{r inference_usage, warning=FALSE}
qf <- hypothesisTest(qf, i = "precursorsPTM_usage", contrast = L, overwrite = TRUE)
```

We extract the results table from the `proteins` summarised experiment in the `qf` object. 

```{r results_usage}
inferencesUsage <- 
  msqrobCollect(qf[["precursorsPTM_usage"]], L)
```

## Report results

We report the results using a results table, volcano plots and heatmaps. 

### Results table

```{r result_tables_usage}
#| results: asis
for (j in colnames(L)) {

  inference <- inferencesUsage |> 
    dplyr::filter(adjPval < alpha & contrast == j)

  cat("**Median - Contrast:**", j, "= 0 (", nrow(inference),
      "significant proteins)\n\n")

  cat('<div style="max-height:300px; overflow-y:auto;">')

  print(
    kable(
      inference |>
        dplyr::arrange(pval) |>
        dplyr::relocate(feature),
      row.names = FALSE
    )
  )

  cat('</div>')

  cat("\n\n\n---\n\n")
}
```

### Volcanoplots 

```{r}
volcanoplotUsage <- plotVolcano(inferencesUsage) + 
  facet_wrap(~contrast) + 
  labs(title = "Usage")
volcanoplotUsage
```


```{r}
inferencesUsage |> 
  filter(adjPval < alpha) |> 
  mutate(DA = sign(logFC) |> as.factor() |> recode("-1"= "down","1" = "up")) |>
  group_by(contrast, DA) |> 
  ggplot(aes(x = contrast)) +
  geom_bar(aes(fill = factor(DA)),
           colour = "black") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

We observe that the number of significant precursors reduced drastically. 

### Heatmaps

```{r heatmaps_usage, eval=FALSE}
heatmaps <- lapply(colnames(L),
                   function(contrast, se, alpha)
                   {
                     sig <- rowData(se)[[contrast]] |> 
                       filter(adjPval < alpha) |> 
                       rownames()
                     quants <- t(scale(t(assay(se[sig,]))))
                     colnames(quants) <- se$sampleId #specific to this dataset to get short colnames
                     rowclushlp <- quants
                     rowclushlp[is.na(rowclushlp)] <- min(quants,na.rm=TRUE) - 2
                     rowclus <- hclust(dist(rowclushlp))
                     annotations <- columnAnnotation(
                       group = se$strain
                       ) #3.
                     set.seed(1234) ## annotation colours are randomly generated by default
                     return(
                       Heatmap(
                       quants, 
                       name = "log2 intensity",
                       top_annotation = annotations, 
                       column_title = paste0(contrast, " = 0"),
                       cluster_rows = rowclus
                       )
                     )
                   },
                   se = getWithColData(qf, "precursorsPTM_usage"),
                   alpha = alpha)
heatmaps
```

### Detail plots

We can explore the data for a protein to validate the statistical inference results. For example, let’s explore the normalised precursor and the summarised protein intensities for the protein with the most significant log2 fold change.

```{r}
(target_feature <- inferencesUsage |> 
   dplyr::slice(which.min(pval)) |> 
   pull(feature)
)
inferencesUsage |> 
  filter(feature == target_feature)
```

To obtain the required data, we perform a little data manipulation pipeline:

We use the QFeatures subsetting functionality to retrieve all data related to and focusing on the peptides_log and proteins sets that contains the peptide ion data used for model fitting. We then convert the data with longForm() for plotting. Finally, we plot the log2 normalised intensities for each sample at the protein and at the peptide level. Since multiple peptides are recorded for the protein, we link peptides across samples using a grey line. Samples are colored according to E. coli spike-in condition.

```{r}
ptm <-  qf[target_feature, , c("precursorsPTM_norm", "precursorsPTM_usage")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()
protein <- qf[unique(ptm$Protein.Group), , c("proteins")] |> #1
  longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |> #2
  data.frame()


ptm_protein <- rbind(ptm, protein)
ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))

outplot <- ptm_protein |>
  ggplot() +
  aes(x = colname,
      y = value) +
  geom_line(aes(group = rowname), linewidth = 0.1) +
  geom_point(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) +
  labs(subtitle = target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
print(outplot)
```

```{r}
rbind(ptm, protein) %>%
  {
  ggplot(.) +
  aes(x = strain,
      y = value) +
  geom_boxplot(aes(colour = strain)) +
  facet_wrap(~ assay, scales = "free") +
  geom_jitter(aes(shape = rowname)) +
  scale_shape_manual(values = seq_len(dplyr::n_distinct(.$rowname))) +
  ggtitle(target_feature) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  guides(shape = "none")
  }
```

We compare the intensities for the top 5 DA phospo-precursors for the first 
contrast to these of their corresponding protein in the non-enriched assay and
to the usages. 
 
```{r comparison_plots_top5, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesPTM |> 
  filter(contrast == contr) |> 
  arrange(pval) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm","precursorsPTM_usage")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3),
                       "padj usage =", 
                       inferencesUsage |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |>
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) 
 print(comparison_plot)
}
```

We compare the intensities for the top 5 DU phospo-precursors for the first 
contrast to these of their corresponding protein in the non-enriched assay and
to their unadjusted abundances. 
 
```{r comparison_plots_top5_DU, warning = FALSE, message = FALSE}
contr = colnames(L)[1]
top5 <- inferencesUsage |> 
  filter(contrast == contr) |> 
  arrange(pval) |> 
  head(n = 5) |> 
  pull(feature)

for (feat in top5)
{
  ptm_data <- qf[,,c("precursorsPTM_norm","precursorsPTM_usage")] |> 
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(rowname==feat)
  feature_protein <- ptm_data |> 
    pull("Protein.Group") |> 
    unique()
  prot_data <- qf[,,"proteins"]|>
    longForm(colvars = colnames(colData(qf)), rowvars = "Protein.Group") |>
    data.frame() |> 
    filter(Protein.Group==feature_protein) 
  ptm_protein <- rbind(ptm_data, prot_data)
  ylims <- ptm_protein |> 
  group_by(assay) |> 
  summarise(cent = mean(range(value,na.rm=TRUE)), ampl = diff(range(value,na.rm=TRUE))) |> 
  mutate(lower = cent - max(ampl)/2, 
         upper = cent + max(ampl)/2) |> 
  select(-c(cent, ampl))
  
  comparison_plot <- ptm_protein |>
    ggplot() +
    aes(x = sampleId,
        y = value) +
    geom_line(aes(group = rowname), linewidth = 0.1) +
    geom_point(aes(colour = strain)) +
    facet_wrap(~ assay, scales = "free") +
    labs(
      title = paste0(feat," / ", feature_protein),
      subtitle = paste("padj PTM =", 
                       inferencesPTM |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |> 
                         round(digits = 3),
                       "padj usage =", 
                       inferencesUsage |> 
                         filter(feature == feat & contrast == contr) |> 
                         pull(adjPval) |>
                         round(digits = 3))
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    ggh4x::facetted_pos_scales(
    y = list(
      assay == ylims$assay[1] ~ scale_y_continuous(limits = unlist(ylims[1,c("lower","upper")])),
      assay == ylims$assay[2] ~ scale_y_continuous(limits = unlist(ylims[2,c("lower","upper")])),
      assay == ylims$assay[3] ~ scale_y_continuous(limits = unlist(ylims[3,c("lower","upper")]))
    )
    ) 
 print(comparison_plot)
}
```


# Differential usage analysis at the PTM level

Multiple precursors are carrying the same PTM (cover the PTM at a certain 
position in the protein). 
We can aggregate them together. 

The variable Protein.Site in the DIA-NN parquet file contains information on the 
exact modified residue position in the full protein sequence. 

```{r}
     rowData(qf[["precursorsPTM_usage"]])$mods <- str_extract_all(rowData(qf[["precursorsPTM_usage"]])$Modified.Sequence, "UniMod:[0-9]+") |> sapply(function(x) str_remove_all(paste(x, collapse = ","), "UniMod:"))
     rowData(qf[["precursorsPTM_usage"]])$sites = str_extract(rowData(qf[["precursorsPTM_usage"]])$Protein.Sites, "(?<=:)[^\\]]+")
```
